D-Hammer: Efficient Equational Reasoning for Labelled Dirac Notation

### 2025-01-31 05tool_eval.tex
update

\section{Implementation and Case Study}
The improvement and extensions above concludes in our implementation called D-Hammer, a solver written in \CC. It has a parser built by ANTLR4, and scalar reasonings are powered by a Mathematica kernel. The user can use commands to make definitions and assumptions in the maintained context, conduct the normalization and equivalence checking, and obtain the rewriting trace output. 
It can be used from the command line interactively, or can be integrated into other \CC\ projects as a library.
This implementation is tested on the benchmark of the previous work and new examples for labelled Dirac notation, demonstrating significant improvement in expressivity and efficiency. 
The tool and example encodings will be open-source and public available.

\subsubsection{Project Structure}
The project structure is illustrated in~\Cref{fig: dhammer structure}.
\texttt{ualg} is the module for universal algebra, defining basic concepts like terms and substitutions. It serves as the library for \texttt{dhammer}, which are then utilized in the example benchmarks and the toplevel command line application. The components of \texttt{dhammer} are as follows:
\begin{itemize}
    \item \texttt{symbols.cpp}: the reserved symbols and AC symbols;
    \item \texttt{syntax\_theory.cpp}: syntax related algorithms, such as de Bruijn normalization and freeness of variables;
    \item \texttt{calculus.cpp}: type checker and intergration with Mathematica;
    \item \texttt{reduction.cpp}: all the rewriting rules and transformations;
    \item \texttt{dirac\_parser.cpp}: parser for Dirac notation and D-Hammer commands;
    \item \texttt{prover.cpp}: the prover that maintains the context and process commands like definition or equivalence checking.
\end{itemize}

\input{fig/proj_structure.tex}

The internal data structure for terms is a pointer-based syntax tree following the function application style:
\[
    \texttt{
        term ::= ID | ID [term (, term)*].
    }
\]
The syntax tree can be an identifier, or an application with an identifier as the function head, and several syntax trees as arguments. There are several Dirac notation terms and their corresponding syntax trees.
\footnotesize{
\begin{align*}
    & X_1 + X_2 + X_3 && \texttt{ADD[X1, X2, X3]} 
    \\
    & \lambda x: \OType(T_1,  T_2). x^\dagger && \texttt{FUN[x, OTYPE[T1, T2], ADJ[x]]}
    \\
    & \sum_{i \in \mathbf{U}(T)} \ket{i} \bra{i} && \texttt{SUM[USET[T], FUN[i, BASIS[T], OUTER[KET[i], BRA[i]]]]}
\end{align*}
}
The syntax tree structure is also compatible with the datatype of Mathematica. This improves the interoperability between D-Hammer and the Mathematica system, enabling them to work interleavingly.
To improve usability, D-Hammer also supports many special notations for terms,  and most Dirac notation terms will be encoded in the natural way.
Here are some examples for the parsing syntax.

@@

Finally, D-Hammer uses a prover to host the computation. The prover maintains a well-formed context $E$, and processes commands to modify the context and conduct calculations. The commands are listed below.
\begin{itemize}
    \item \texttt{\textcolor{NavyBlue}{Def} ID := term.} It defines the \texttt{ID} as the \texttt{term}, using the \textbf{W-Def} typing rule.
    \item \texttt{\textcolor{NavyBlue}{Var} ID := term.} It make an assumption of \texttt{ID} with the \texttt{term} as type, using the \textbf{W-Assume} typing rules.
    \item \texttt{\textcolor{NavyBlue}{Check} term.} Type checking the \texttt{term} and output the result.
    \item \texttt{\textcolor{NavyBlue}{Normalize} term.} Normalize the \texttt{term} using the algorithm introduced in~\Cref{sec: decide}.
    \item \texttt{\textcolor{NavyBlue}{CheckEq} term \textcolor{NavyBlue}{with} term.} Check the equivalence of the two terms calculating and comparing their normal forms.
\end{itemize}
The prover will type check the terms for each command. We can also use \texttt{\textcolor{NavyBlue}{Normalize} term \textcolor{NavyBlue}{with trace}.} to output the proof trace during normalization. The proof trace is a sequence of records, including the rule or transformation appied, the position of application, and the pre- and post-transformation terms. The record helps understand the normalization procedure better, and can be turned into verified proofs in theorem provers in the future.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\textwidth]{fig/coqq.pdf}
    \caption{Time comparison between DiracDec and D-Hammer on the CoqQ benchmark.}
    \label{fig: CoqQ plot}
\end{figure}

@@

\subsubsection{Labelled Dirac Notation}
We present a new set of examples for labelled Dirac notation (LDN), as detailed in~\Cref{sec: examples for labelled}. These examples, illustrated in~\Cref{fig: labelled examples}, include six representative cases drawn from a variety of sources, such as well-established theorems, research paper results, quantum circuit equivalence, and quantum program verification.
D-Hammer normalizes them and successfully checks their equivalence using the algorithm in~\Cref{sec: labelled}.
Among the examples, LDN-12 shows the flexibility in combining labelled Dirac notations.
Example LDN-14 and LDN-18 are about circuits or programs on qubits. LDN-14 shows how to calculate controlled-not gate in different ways. In LDN-18, $q, q_1$ and $q_2$ represent variables for quantum programs, and the equation shows the change of the state after applying $X$ gate on $q_2$. Experiments show that our algorithm is relatively inefficient for labelled Dirac notations based on qubits.

@@

\begin{figure}[h]
    \center
    % \scriptsize
    \setlength{\extrarowheight}{2pt}
    \begin{tabular}{l c c l}
        \hline
        example & source & time(s) & equation \\
        \hline
        LDN-4 & theorem & 0.03 & \( \bra{\Phi}_{(x, y)} M_y \cdot N_x \ket{\Phi}_{(x, y)} = \mathrm{tr}(M^T N) \) \\
        LDN-10 & paper~\cite{DBLP:conf/lics/ZhouBHYY21} & 5.17 & \( \mathrm{tr}_{((a',(b,b')),c')}\Big[\mathrm{tr}_{r}\Big(U_{(r,(a,b))} \cdot\Big(|s\>_r\<s|\otimes 
        \Big[V_{((a',(b,b')),c')} \cdots \)\\
        LDN-11 & paper~\cite{PALSBERG2024206} & 0.12 & \( U_{(a,b)}\cdot W_{(b,c)}\cdot V_{(a,c)} = 
        \sum_i |i\>_a\<i|\otimes \big((P_i)_c\cdot W_{(b,c)}\cdot (Q_i)_c\big)\) \\
        LDN-12 & circuit & 0.07 & \( \ket{i}_{a;b}\bra{j} \cdot C_{(b,c)} \cdot D_{(c,d)} = {}_b\bra{j} \cdot C_{(b,c)} \cdot D_{(c,d)} \cdot \ket{i}_{a}\) \\
        LDN-14 & circuit & 7.37 & \scriptsize{\( \textsf{CNOT}_{rq}\ket{\textrm{GHZ}}_{pqr} = (\textsf{CNOT}\ket{00})_{rq}\ket{0}_p + (\textsf{CNOT}\ket{11})_{rq}\ket{1}_p \)} \\
        LDN-18 & program & 1.01 & \( -\ket{0}_q\ket{+-}_{q_1, q_2} = X_{q_2} \ket{0}_q\ket{+-}_{q_1, q_2} \) \\
        \hline
    \end{tabular}
    \caption{Part of examples for labelled Dirac notations.}
    \label{fig: labelled examples}
\end{figure}

>>>>

\section{Implementation and Case Study}
The improvement and extensions above concludes in our implementation called D-Hammer, a solver written in \CC. It has a parser built by ANTLR4, and scalar reasonings are powered by Mathematica Engine. The user can use commands to make definitions and assumptions in the maintained context, conduct the normalization and equivalence checking, and obtain the rewriting trace output. 
It can be used from the command line interactively, or can be integrated into other \CC\ projects as a library.
This implementation is tested on the benchmark of the previous work and new examples for labelled Dirac notation, demonstrating significant improvement in expressivity and efficiency. 
The tool and example encodings will be open-source and publicly available.

\subsubsection{Structure and Mechanism} 
The project consists of the following components:
\begin{itemize}
    \item \texttt{antlr4}: the third-party library for building the parser.
    \item \texttt{WSTP interface}: the wrapper to link with Mathematica Engine.
    \item \texttt{ualg}: the framework module for universal algebra, defining basic concepts like terms and substitutions.
    \item \texttt{dhammer}: the main module containing symbols definitions, type checking, rewriting rules, normalization algorithm and the prover.
    \item \texttt{example}: an example benchmark for evaluation.
    \item \texttt{toplevel}: the command line application.
\end{itemize}
% The project structure is illustrated in~\Cref{fig: dhammer structure}.
% \texttt{ualg} is the module for universal algebra, defining basic concepts like terms and substitutions. It serves as the library for \texttt{dhammer}, which are then utilized in the example benchmarks and the toplevel command line application. Important components of \texttt{dhammer} are:

% \input{fig/proj_structure.tex}

The internal data structure for terms is a pointer-based syntax tree following the function application style:
\[
    \texttt{
        term ::= ID | ID [term (, term)*].
    }
\]
The syntax tree can be an identifier, or an application with an identifier as the function head, and several syntax trees as arguments. There are several Dirac notation terms and their corresponding syntax trees.
\footnotesize{
\begin{align*}
    & X_1 + X_2 + X_3 && \texttt{ADD[X1, X2, X3]} 
    \\
    & \lambda x: \OType(T_1,  T_2). x^\dagger && \texttt{FUN[x, OTYPE[T1, T2], ADJ[x]]}
    \\
    & \sum_{i \in \mathbf{U}(T)} \ket{i} \bra{i} && \texttt{SUM[USET[T], FUN[i, BASIS[T], OUTER[KET[i], BRA[i]]]]}
\end{align*}
}
% The syntax tree structure is also compatible with the datatype of Mathematica. This improves the interoperability between D-Hammer and the Mathematica system, enabling them to work interleavingly.
To improve usability, D-Hammer also supports many special notations for terms,  and most Dirac notation terms will be encoded in the natural way.
Here are some examples for the parsing syntax.

@@

Finally, D-Hammer uses a prover to host the computation. The prover maintains a well-formed context $\Gamma$, and processes commands to modify the context and conduct calculations. The commands are listed below.
\begin{itemize}
    \item \texttt{\textcolor{NavyBlue}{Def} ID := term.} It defines the \texttt{ID} as the \texttt{term}, using the \textbf{W-Def} typing rule.
    \item \texttt{\textcolor{NavyBlue}{Var} ID := term.} It make an assumption of \texttt{ID} with the \texttt{term} as type, using the \textbf{W-Assume} typing rules.
    \item \texttt{\textcolor{NavyBlue}{Check} term.} Type checking the \texttt{term} and output the result.
    \item \texttt{\textcolor{NavyBlue}{Normalize} term.} Normalize the \texttt{term} using the algorithm introduced in~\Cref{sec: decide}.
    \item \texttt{\textcolor{NavyBlue}{CheckEq} term \textcolor{NavyBlue}{with} term.} Check the equivalence of the two terms calculating and comparing their normal forms.
\end{itemize}
The prover will type check the terms for each command. We can also use \texttt{\textcolor{NavyBlue}{Normalize} term \textcolor{NavyBlue}{with trace}.} to output the proof trace during normalization. The proof trace is a sequence of records, including the rule or transformation appied, the position of application, and the pre- and post-transformation terms. The record helps understand the normalization procedure better, and can be turned into verified proofs in theorem provers in the future.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.65\textwidth]{fig/coqq.pdf}
    \caption{Time comparison between DiracDec and D-Hammer on the CoqQ benchmark.}
    \label{fig: CoqQ plot}
\end{figure}

@@

\subsubsection{Labelled Dirac Notation}
We present a new set of examples for labelled Dirac notation (LDN). These examples, illustrated in~\Cref{fig: labelled examples}, include six representative cases drawn from a variety of sources, such as well-established theorems, research paper results, quantum circuit equivalence, and quantum program verification.
D-Hammer normalizes them and successfully checks their equivalence using the algorithm in~\Cref{sec: labelled}.
Among the examples, LDN-12 shows the flexibility in combining labelled Dirac notations.
Example LDN-14 and LDN-18 are about circuits or programs on qubits. LDN-14 shows how to calculate controlled-not gate in different ways. In LDN-18, $q, q_1$ and $q_2$ represent variables for quantum programs, and the equation shows the change of the state after applying $X$ gate on $q_2$. 
% Experiments show that our algorithm is relatively inefficient for labelled Dirac notations based on qubits.

@@

\begin{figure}[h]
    \center
    % \scriptsize
    \setlength{\extrarowheight}{2pt}
    \begin{tabular}{l c c l}
        \hline
        example & source & time(s) & equation \\
        \hline
        LDN-4 & theorem & 0.03 & \( \bra{\Phi}_{(x, y)} M_y \cdot N_x \ket{\Phi}_{(x, y)} = \mathrm{tr}(M^T N) \) \\
        LDN-10 & paper~\cite{DBLP:conf/lics/ZhouBHYY21} & 5.17 & \( \mathrm{tr}_{((a',(b,b')),c')}\Big[\mathrm{tr}_{r}\Big(U_{(r,(a,b))} \cdot\Big(|s\>_r\<s|\otimes 
        \Big[V_{((a',(b,b')),c')} \cdots \)\\
        LDN-11 & paper~\cite{PALSBERG2024206} & 0.12 & \( U_{(a,b)}\cdot W_{(b,c)}\cdot V_{(a,c)} = 
        \sum_i |i\>_a\<i|\otimes \big((P_i)_c\cdot W_{(b,c)}\cdot (Q_i)_c\big)\) \\
        LDN-12 & circuit & 0.07 & \( \ket{i}_{a;b}\bra{j} \cdot C_{(b,c)} \cdot D_{(c,d)} = {}_b\bra{j} \cdot C_{(b,c)} \cdot D_{(c,d)} \cdot \ket{i}_{a}\) \\
        LDN-14 & circuit & 7.37 & \scriptsize{\( \textsf{CNOT}_{rq}\ket{\textrm{GHZ}}_{pqr} = (\textsf{CNOT}\ket{00})_{rq}\ket{0}_p + (\textsf{CNOT}\ket{11})_{rq}\ket{1}_p \)} \\
        LDN-18 & program & 1.01 & \( -\ket{0}_q\ket{+-}_{q_1, q_2} = X_{q_2} \ket{0}_q\ket{+-}_{q_1, q_2} \) \\
        \hline
    \end{tabular}
    \caption{Part of examples for labelled Dirac notations. See~\Cref{sec: examples for labelled} for the full list.}
    \label{fig: labelled examples}
\end{figure}

### 2025-01-31 06related_conclusion.tex
update

\section{Conclusion and Future Work}
We have designed and implemented D-Hammer, a dependently typed
higher-order language and proof system for labelled Dirac
notation. D-Hammer benefits from an optimized implementation in
\CC\ and a tight integration with Mathematica to reason about a broad
range of mathematical functions, including trigonometric and
exponential functions, that are commonly used in quantum physics.
There are three important directions for future work. The first
direction is to enhance D-Hammer with support for big tensor notation,
i.e.\, to support expressions of the form $\bigotimes_{i\in I}
e_i$. The second direction is to extend D-Hammer with a mechanism to
generate independently verifiable certificates.  There is a large body
of work on producing certificates for automated tools, in particular
SMT solvers; see e.g.~\cite{DBLP:journals/cacm/BarbosaBCDKLNNOPRTZ23}
for a recent overview. One potential option would be to integrate
D-Hammer with the Coq or Lean proof assistants; in the first case, one
would benefit from the formalization of labelled Dirac notation in
CoqQ~\cite{Zhou2023}, whereas in the second case, one would benefit
from powerful mechanisms to integrate rewriting procedures into the
Lean proof assistant. Finally, the third direction is to connect
D-Hammer with quantum program verifiers. Two potential applications
are automating equational proofs for tools that already use Dirac
notation, and to substitute numerical methods for tools that use
matrices instead of symbolic assertions.

>>>>

\section{Conclusion and Future Work}
We have designed and implemented D-Hammer, a dependently typed
higher-order language and proof system for labelled Dirac
notation. D-Hammer benefits from an optimized implementation in
\CC\ and a tight integration with Mathematica to reason about a broad
range of mathematical functions, including trigonometric and
exponential functions, that are commonly used in quantum physics.
There are two important directions for future work. 
% The first
% direction is to enhance D-Hammer with support for big tensor notation,
% i.e.\, to support expressions of the form $\bigotimes_{i\in I}
% e_i$. 
The first direction is to extend D-Hammer with a mechanism to
generate independently verifiable certificates.  There is a large body
of work on producing certificates for automated tools, in particular
SMT solvers; see e.g.~\cite{DBLP:journals/cacm/BarbosaBCDKLNNOPRTZ23}
for a recent overview. One potential option would be to integrate
D-Hammer with the Coq or Lean proof assistants; in the first case, one
would benefit from the formalization of labelled Dirac notation in
CoqQ~\cite{Zhou2023}, whereas in the second case, one would benefit
from powerful mechanisms to integrate rewriting procedures into the
Lean proof assistant. The second direction is to connect
D-Hammer with quantum program verifiers. Two potential applications
are automating equational proofs for tools that already use Dirac
notation, and to substitute numerical methods for tools that use
matrices instead of symbolic assertions.
